---
title: "Predict Microbial Community"
substitle: "Script 2: Modelling soil properties and fungal communities"
author: Natacha Bodenhausen
date: '`r Sys.Date()`'
output:
  pdf_document:
        toc: yes
urlcolor: blue
---


```{r setup, include=FALSE}


#libraries
knitr::opts_chunk$set(echo = FALSE, 
                      cache = FALSE)


suppressPackageStartupMessages(library(tidyverse))
library(vegan) # not sure if it necessary because it is in phyloseq
library("phyloseq")
library("kableExtra") # to style the tables 
library(soiltexture) # ternary plot
library(ggcorrplot) # correlogram
#library("factoextra") # plot PCA results
# library(gplots) #heatmap.2
library(gridExtra) ## grid_arrange
library(cowplot) # plot_grid => several plots in one figure
library(viridis) # color for people with colorblindness  
library(ggrepel) # geom_text_repel


# same color as ggplot
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

set.seed(8037)

# logical: would you like figures?
makeFig <- TRUE

# logical: we would you like to work with first sequencing  
# where each year was sequenced separately or not (m11)
m1m8m9 <- TRUE

```


```{r}

my_theme <- theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

  
theme_main <-theme(text = 
                     element_text(size = 10))

theme_supp <-theme(text = 
                     element_text(size = 10))

theme_presentation <-theme(text = element_text(size = 20))



```

\newpage

# Material and methods 

## 2018

22 fields were sampled in the spring 2018

4 DNA extractions per field done by Alain in the summer 2018.

ITS1F and ITS4 primers were used to amplify both ITS1 and ITS2.

Library "microbial1" was prepared by Alain at Reckenholz and sequenced by FGCZ.


## 2019

25 fields were sampled in the spring 2019

4 DNA extractions per field done by Alain in the spring 2020.

ITS1F and ITS4 primers were used to amplify both ITS1 and ITS2.

Library "microbial6" was prepared by Markus at UniBasel and sequenced by Unibern.

Quality of sequencing really bad, so Gabriel prepared a second library, called "microbial8".

## 2020

12 fields were sampled in the spring 2020

4 DNA extractions per field  done by Alain in the winter 2021.

ITS1F and ITS4 primers were used to amplify both ITS1 and ITS2.

Library "microbial9" was prepared by Gabriel at Uni Basel and sequenced in Bern.

## 2022

All samples were resequenced in a fourth library "microbial11" in 2022. The four DNA replicates were pooled before the PCR, so there is only one PCR per field.

## Bioinformatics

Prepared by Jan Wälchli with new taxonomy from November 2021.

```{r}
print("This script uses either data from the three librairies (m1m8m9) or fourth library(m11)")
```

```{r, echo = TRUE}
m1m8m9

```


\newpage

```{r}


if(m1m8m9 == TRUE){
  load(file="../results/physeq_m1_m8_m9_norm.rda")
} else {
   load(file="../results/physeq_m11_norm.rda")
  }

load(file="../results/soil_selected.rda")

# add soil variables to the phyloseq object
design_table <- data.frame(sample_data(physeq_norm))

design_table <-left_join(design_table,
                         soil_selected, by ="field") %>%
  tibble::column_to_rownames("field")

sample_data(physeq_norm) <- design_table

# we need a variable field
sample_data(physeq_norm)$field <- sample_names(physeq_norm)



```

```{r}

env_data <-data.frame(sample_data(physeq_norm))

# for the RDA
# reduced_set<- env_data %>% dplyr::select(-field,- library, - year)
reduced_set<- env_data %>% dplyr::select(Mg:Na)


# OTU
taxonTable <-otu_table(physeq_norm, taxa_are_rows = FALSE)
taxon_table_hell <- sqrt(taxonTable) # hellinger is sqrt relative abundance

```


\newpage

# Constrained Analysis

## forward selection with ordistep

```{r, echo=FALSE}

mod0<- rda(taxon_table_hell ~ Condition(env_data$library) , data = reduced_set)

mod1 <- rda(taxon_table_hell ~ .+Condition(env_data$library), data = reduced_set)

anova(mod1)
```

```{r, eval = FALSE}

# takes a lot of time 
# only run this code chunk by hand

 mod2 <-ordistep(mod0, scope = formula(mod1),  direction = 'forward', permutations = 999)

 
if(m1m8m9 == TRUE) {
   save(mod2,file="../results/ordistep.rda")}else {
   save(mod2,file="../results/ordistep_m11.rda")
 }


p.adjust(mod2$anova$`Pr(>F)`, method= "holm")

mod2$call
```

rda(formula = taxon_table_hell ~ Condition(env_data$library) + 
    pH + K + sand + MIC + B + Nmin + resp + Mn + clay + Ptot, 
    data = reduced_set)
    
Different model with microbials11 (second sequencing), iron is new and clay and Ptot are missing

Call: rda(formula = taxon_table_hell ~ Condition(env_data$library) +
pH + sand + K + MIC + B + resp + Fe + Mn + Nmin, data = reduced_set)


## forward selection with ordiR2step


```{r, echo = FALSE, eval = FALSE}
# takes a bit less time than ordistep
# only run this code chunk by hand

adjR2.tbrda <- RsquareAdj(mod1)$adj.r.squared #   0.1560872 - adjusted R2 explained by all variables is 16%

  mod3<- ordiR2step (mod0, scope = formula (mod1), R2scope = adjR2.tbrda, direction = 'forward', permutations = 999)
#  
# save(mod3, file="../results/ordi2step.rda")

load(file="../results/ordi2step.rda")


mod3$anova
mod3$call


 # write.csv(mod3$anova, "../results/ordi2step_results.csv.csv")


#p.adjust(mod3$anova$`Pr(>F)`, method= "holm")

global_anova <-anova(mod3, permutations=how(nperm=999))

RsquareAdj(mod3)$adj.r.squared
```

rda(formula = taxon_table_hell ~ Condition(env_data$library) + 
    pH + K + clay + MIC + resp + Nmin +  Mn +B + Ptot + sand , 
    data = reduced_set)

Comparing model from ordistep and from  ordiR2step: they are a bit different, WHC is only present with the first call, Ptot only present with the second call.

Since the results are similar but ordiR2step is faster, let's use the second one for the leave-one-out validation.


\newpage

## RDA with only the variables selected by ordiR2step with m1_m8_m9


```{r, echo = TRUE}


var_model<- env_data %>% 
  dplyr::select(field, library, pH, K,  clay,  MIC, resp, Nmin, Mn,B,  Ptot, sand)

variable_selection <- var_model %>%
  dplyr::select(-library, -field)

```


```{r, repeat_rda}


mod4 <- rda(taxon_table_hell ~ .+ Condition(var_model$library), data = variable_selection)

mod4$call



#summary(mod4, axes=0)
print("R2")
(R2<-RsquareAdj(mod4)$r.squared)
print("adjusted R2")
(R2<-RsquareAdj(mod4)$adj.r.squared)

# takes a lot of time
# p 221: global test of the RDA results

 # global_anova <-anova(mod4, permutations=how(nperm=999))
# save(global_anova , file = "../results/global_anova.rda")


# test of all canonical axes 
# takes a lot of time

 # result_anova_axis <-anova(mod4, by="axis", permutations=how(nperm=999))
 # save(result_anova_axis , file = "../results/anova_axis.rda")

load(file="../results/anova_axis.rda")
load(file="../results/global_anova.rda")

print("global test")

global_anova

print("test all canonical axes")

result_anova_axis

# the first 4 axes are significant

# Borcard et al p 213 and 216
#
# plot(mod4, scaling=2, display=c("lc","cn"))
# they recommend to use the lc scores instead of the wa

# axes three and four
# plot(mod4, scaling=2, display=c("lc","cn"), choices = c(3,4))

```

Klaus's question: what is R2 if not correcting for year/library?

```{r}

mod5 <- rda(taxon_table_hell ~ ., data = variable_selection)

print("R2")
(R2<-RsquareAdj(mod5)$adj.r.squared)

```

\newpage


## Triplot RDA 



```{r prepare_data_plot}

#ord$eig
eig2 <- eigenvals(mod4)
#eig2 / sum(eig2)
#cumsum(eig2 / sum(eig2))

percent_RDA <-round(eig2 / sum(eig2)*100,1)

coordi_RDA <- (paste("RDA", c(1:9),
                 paste(" (",percent_RDA[1:9],"%",")",sep=""),
                 sep=""))


species_scores_rda <-scores(mod4, 
                 choice=1:2,
                 scaling="species",
                 display="sp")

sites_scores_rda <-scores(mod4, 
                 choice=1:2,
                 scaling="species",
                 display="lc")


load(file="../results/m1_m8_m9_top_OTU.rda")

# core community
core_species_rda<- which(colnames(taxonTable) %in% top_OTU$otu)

env_arrow_rda <- mod4$CCA$biplot[,1:2]

# save(sites_scores_rda, species_scores_rda, core_species_rda, env_arrow_rda, coordi_RDA, file="../results/triplot_mod4.rda")

```

Interpreting the results of the RDA using the length of the arrows
 
 
```{r, echo = TRUE}
sqrt(rowSums(env_arrow_rda^2) %>% sort())
```


```{r, prepare_data_plot_RDA}

# Settings
ahead <- 0.05   # Length of arrow heads
offset <- 0.1 # distance between arrow head and center of blue text
mult_arrow<-4
mult_species <-4

sites <- left_join(sites_scores_rda %>% 
                     as.data.frame() %>%
                     tibble::rownames_to_column("field"),
                   env_data %>% 
                     dplyr::select(year, field), 
                   by = "field") %>% 
  mutate(year = as_factor(year))

species <- data.frame()

# p 219 Borcard
spe.good.rda <- goodness(mod4) # by default species
sel.sp.rda <- which (spe.good.rda[,2]>=0.3) # cumulative


species <-  (species_scores_rda * mult_species) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("OTU") %>% 
  #mutate(has_label = OTU %in% top_OTU$otu)
  mutate(has_label = OTU %in% names(sel.sp.rda))

# alternatively label extreme sites
species_label <- species %>% 
  filter(abs(RDA1)>1 | abs(RDA2)>1)



environment <- (env_arrow_rda * mult_arrow) %>%
  as.data.frame() %>% 
  dplyr::select("RDA1", "RDA2") %>%
  tibble::rownames_to_column("environment") %>%
  mutate(len = sqrt(RDA1^2 + RDA2^2), 
       hoffset = RDA1/len*offset, voffset = RDA2/len*offset)

```



```{r plot_triplot_ggplot2}

# Plot
fig <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(aes(RDA1, RDA2), color = "grey", 
             shape = "cross", data = species) +
  geom_text(aes(RDA1, RDA2, label = OTU), color = "grey", 
              hjust = 0, nudge_x = .05, 
            data = species %>% filter(.data$has_label)) +
  geom_point(aes(RDA1, RDA2, color = year), data = sites) +
  geom_text(aes(RDA1, RDA2, color = year, label = field), 
              hjust = 0, nudge_x = .05, data = sites) +
  geom_segment(aes(x = 0, y = 0, xend = RDA1, yend = RDA2), 
                 color = "blue", 
               arrow = arrow(length=unit(0.02,"npc"), type = "open"),
               data = environment) +
  geom_text(aes(RDA1 + hoffset, RDA2 + voffset, 
                label = environment), color = "blue",
            data = environment) +
  coord_equal(ratio = 1, clip = "off") +
  theme_bw() +
  labs(x = coordi_RDA[1], y = coordi_RDA[2]) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.position = "none")

print(fig) 
  
b <- ggplot_build(fig)
b$layout$panel_params[[1]]$x.range
b$layout$panel_params[[1]]$y.range

if(makeFig == TRUE)
  ggsave("../figures/RDA_triplot.pdf",
      width = 12, height = 12, units = "cm")

```

## two biplots


```{r plot_biplot_1, fig.height = 4}

# https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control/25062509
geom.text.size = 3
theme.size = (14/5) * geom.text.size


# Plot
fig1 <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  # geom_point( data = sites,
  #             aes(RDA1, RDA2, color = year)) +
  geom_text(data = sites,
            aes(RDA1, RDA2, color = year, label = field), 
              #vjust = 0.2, nudge_y = .1,  
            size = geom.text.size) +
  geom_segment(data = environment,
               aes(x = 0, y = 0, 
                   xend = RDA1, yend = RDA2), 
                 color = "blue", 
               arrow = arrow(length=unit(0.02,"npc"), 
                             type = "open")) +
  geom_text(data = environment, 
            aes(RDA1 + hoffset, RDA2 + voffset, 
                label = environment), 
            size = geom.text.size,
            color = "blue") +
    labs(x = coordi_RDA[1], y = coordi_RDA[2]) +
   xlim(-3.3,2.6)+
  #ylim(-3,3)+
   coord_equal(ratio = 1, clip = "off") +
  theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.position = "none")

fig1

```



```{r plot_biplot_2, fig.height = 4}

# Plot
fig2 <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(data = species,
             aes(RDA1, RDA2), 
             color = grey(0.4), size = 1)  +
  geom_text(data = species  %>% 
              filter(abs(RDA1)>1 | abs(RDA2)>1),
            aes(RDA1, RDA2, label = OTU), 
            color = grey(0.4), 
            hjust = 0.2, nudge_y = .2, 
            size = geom.text.size) +
   geom_segment(data = environment,
                aes(x = 0, y = 0, 
                    xend = RDA1, yend = RDA2), 
                 color = "blue", 
               arrow = arrow(length=unit(0.02,"npc"), 
                             type = "open")) +
  geom_text(data = environment,
            aes(RDA1 + hoffset, RDA2 + voffset, 
                label = environment),
            size = geom.text.size, 
            color = "blue") +
   xlim(-3.3,2.6)+
   #ylim(-3,3)+
    coord_equal(ratio = 1, clip = "off") +
  labs(x = coordi_RDA[1], y = coordi_RDA[2]) +
  theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(legend.position = "none")

fig2

```

```{r}

two_biplots <-plot_grid(fig1 + theme_main, 
                        fig2 + theme_main, 
                        labels = c('A', 'B'),
                        label_size = 12)

print(two_biplots)

if(makeFig == TRUE)
  ggsave("../figures/rda_both_biplots.pdf",
       two_biplots,
       width = 16, height = 9, units = "cm")

```

 
\newpage

## interpreation of RDA

```{r}

# ggplot(for_gg, aes(x = var, y = otu)) + 
#   geom_point() +
#   stat_smooth(method = "lm", col = "red")


ggplotRegression <- function (fit, otu, var) {
ggplot(fit$model, 
       aes_string(x = names(fit$model)[2], 
                  y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(method = "lm", col = "red") +
  labs(x = var,
       y = "abdundance",
       title = otu,
       subtitle = 
         paste("Adj. R2 = ", signif(summary(fit)$adj.r.squared, 2),
                     " P =" , signif(summary(fit)$coef[2,4], 2)))+
    theme_bw()
  
  
}

```


```{r, echo = FALSE}

xx<-as.data.frame (taxon_table_hell)

## OTU2 and sand
for_gg <- data.frame("var" = var_model$sand, "otu" = xx$OTU2 )


p1 <- ggplotRegression(lm(otu ~ var, dat = for_gg), 
                       otu = "OTU2", 
                       var = "sand")

## OTU2 and Potassium
for_gg <- data.frame("var" = var_model$K, "otu" = xx$OTU2)


p2 <- ggplotRegression(lm(otu ~ var, dat = for_gg), 
                       otu = "OTU2", 
                       var = "K")


## OTU6
for_gg <- data.frame("var" = var_model$pH, "otu" = xx$OTU6)


p3 <- ggplotRegression(lm(otu ~ var, dat = for_gg), 
                       otu = "OTU6", 
                       var = "pH")
## OTU5
for_gg <- data.frame("var" = var_model$pH, 
                     "otu" = xx$OTU5)


p4<- ggplotRegression(lm(otu ~ var, dat = for_gg), 
                       otu = "OTU5", 
                       var = "pH")

```

```{r, eval = TRUE}

four_regression <-plot_grid(p1 + theme_supp,
                      p2+ theme_supp, 
                      p3+ theme_supp,
                      p4+ theme_supp,
                      labels = c('A', 'B', 'C', 'D'),
                      nrow = 2,
                      label_size = 12)

print(four_regression)

if(makeFig == TRUE) ggsave("../figures/four_regressions.pdf",
       four_regression,
       width = 16, height = 16, units = "cm")

```


